---
title: "Road Accidents"
subtitle: Univariate Analysis
date: '2022-05-05'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=TRUE)
setwd("E:/Montreal-Asphalt-Detectives/Rmarkdown")
library(tidyverse)
library(ggpubr)
source('./helpers.R')
accidents <- read.csv('../data/collisions_routieres_road_accidents.csv',
                        header = TRUE)

```

```{r fig.width=8, fig.height=4, warning=FALSE}

grid.arrange(valueBox(nrow(accidents), "Entries"),
             valueBox(ncol(accidents), "Features"),nrow= 1)


```  

# Feature definitions and summaries

## 1. Time variables

**AN**: Year,  
**DT_ACCDN**: Date,  
**HR_ACCDN**: Hour,  
**JR_SEMN_ACCDN**: Day.  
  
```{r fig.width=12, fig.height=12}


accidents %>% ggplot(aes(x=factor(AN))) +
  geom_bar() +
  stat_count( geom="text", aes(label=..count..) ,vjust = -1)+ 
  labs(title="Accidents Per Year", x="Year", y="Number of Accidents") ->an
 

accidents %>%
  ggplot(aes(x=factor(JR_SEMN_ACCDN,
                      levels=c("LU","MA","ME","JE","VE","SA","DI"),
                      labels = c("Monday", "Tuesday", "Wednesday", "Thursday",
                                 "Friday", "Saturday", "Sunday")))) +
  geom_bar() +
  stat_count( geom="text", aes(label=..count..) ,vjust = -1)+ 
  labs(title="Accidents Per Day", x="Day of Week", y="Number of Accidents") ->jr


accidents$HEURE_ACCDN %>%
  lapply(function(x) gsub("Non précisé", NA, strsplit(x,':')[[1]][1])) %>%
  as.numeric() -> accidents$HR_ACCDN

hr_labels = c("12AM", sprintf("%dAM", 1:11), "12PM", sprintf("%dPM", 1:11))
accidents %>%
  drop_na(HR_ACCDN) %>%
  ggplot(aes(x=factor(HR_ACCDN,
                                  levels=seq(0,23,1),
                                  labels= hr_labels))) +
  geom_bar() +
  scale_x_discrete(breaks=hr_labels[seq(1,25,3)])+
  labs(title="Accidents Per Hour", x="Hour", y="Number of Accidents")  ->hr






ggarrange(an, jr, hr,
          ncol = 2, nrow = 2)
           
```  

The percentage of null values for the hour of accidents is 
`r round(table(accidents$HEURE_ACCDN)[['Non précisé']] / nrow(accidents) * 100,2)`
%.  

Unless mentioned, the variables do not have null values.  

----   
   
## 2. Losses and Damagaes  (27 variables)  

We will divide the 27 variables into 3 groups.   
**The first** group contains one variable: "Gravité: Severity". It's a categorical
variable of 5 categories:  

* Deadly: at least one victim died during the 30 days following the accident.  
* Severe injuries: At least one victim was severely injured.  
* Light injuries: One or more victims were lightly injured.  
* Material damage only: Material damage with value exceeding $2000.  
* Inferior material damage: Material damage with value equal to or below $2000.  
   
*Whenever deaths are counted, deaths happening during the 30 days following
the accident are included*

```{r fig.width=12, fig.height=4, warning=FALSE}

accidents %>%
  count(grav=factor(GRAVITE,
         levels = c("Dommages matériels inférieurs au seuil de rapportage",
                    "Dommages matériels seulement", "Grave", "Léger", "Mortel"),
         labels = c('Material dammages only <= $2k', 'Material dammages only > $2k',
                    'Severe injuries', 'Light injuries', 
                    "Deadly"))) %>%
  mutate(pct = prop.table(n)) %>%
  ggplot(aes(x=reorder(grav,desc(pct)), y= n, label = scales::percent(pct))) +
  #geom_bar() +
  coord_flip() +
  geom_col(position='dodge' ) +
  theme(axis.text.y = element_text(face='bold', angle = 0, size=10),
        panel.background = element_blank(),, axis.ticks.y = element_blank()) +
  geom_text(position = position_dodge(width = .9),    # move to center of bars
            vjust = 0.5,
            hjust = -1,
            size = 6)  +
  labs(title="The Severity of Accidents", y="Count", x="") +
  scale_y_continuous(expand = expansion(add=c(0,20000)))
  
```
      
**The second** group breaks down the injuries. We have the following groups:

* NB_VICTIMES_TOTAL: Total number of victims
    + NB_MORTS: Number of dead victims.
    + NB_BLESSES_GRAVES: Number of victims severely injured.
    + NB_BLESSES_LEGERS: Number of victims lightly injured.
* NB_VICTIMES_PIETON: Total number of pedestrian victims.
    + NB_DECES_PIETON: Number of dead pedestrians.
    + NB_BLESSES_PIETON: Number of injured pedestrians (severe or light).
* NB_VICTIMES_MOTO: Total number of motorcyclist victims.
    + NB_DECES_MOTO: Number of dead motorcyclists.
    + NB_BLESSES_MOTO: Number of injured motorcyclists.
* NB_VICTIMES_VELO: Number of cyclist victims.
    + NB_DECES_VELO: Number of dead cyclists.
    + NB_BLESSES_VELO: Number of injured cyclists

```{r fig.width=12, fig.height=8, warning=FALSE}


accidents %>% 
  select(NB_VICTIMES_TOTAL, NB_MORTS, NB_BLESSES_GRAVES, NB_BLESSES_LEGERS,
         NB_VICTIMES_PIETON, NB_DECES_PIETON, NB_BLESSES_PIETON,
         NB_VICTIMES_MOTO, NB_DECES_MOTO, NB_BLESSES_MOTO,
         NB_VICTIMES_VELO, NB_DECES_VELO, NB_BLESSES_VELO) %>%
  
  colSums() %>%  
  as.data.frame  %>% 
  rename( COUNT = ".") %>%
  mutate(FEATURE = c("Total", "Deaths", "Injuries", "Injuries",
                     rep(c("Total", "Deaths", "Injuries"),3))) %>%
  mutate(CATEGORY = c(replicate(4,"Total"),
                      replicate(3, "Pedestrians"),
                      replicate(3, "Motorcyclists"),
                      replicate(3, "Cyclists"))) %>%
  filter( CATEGORY != 'Total') %>%
  ggplot(aes(x=factor(FEATURE,
                      levels = c("Total","Deaths", "Severe",
                                 "Light", "Injuries")),
                      y=COUNT, fill=CATEGORY)) +
  geom_bar(stat = 'identity',position = 'stack')+  
  facet_wrap(. ~ factor(FEATURE,levels=c("Total", "Injuries", "Deaths")),
             scales = "free",nrow = 1) +
  geom_text(aes(label=comma(COUNT)), position=position_stack(vjust = .3))+
  geom_text(aes(label=CATEGORY), position=position_stack(vjust = .7)) +
  theme(legend.position = "none", axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
labs(title="Victims Breakdown", y="", x="") 



accidents %>% 
  select(NB_VICTIMES_TOTAL, NB_MORTS, NB_BLESSES_GRAVES, NB_BLESSES_LEGERS,
         NB_VICTIMES_PIETON, NB_DECES_PIETON, NB_BLESSES_PIETON,
         NB_VICTIMES_MOTO, NB_DECES_MOTO, NB_BLESSES_MOTO,
         NB_VICTIMES_VELO, NB_DECES_VELO, NB_BLESSES_VELO) %>%
  
  colSums() %>%  
  as.data.frame  %>% 
  rename( COUNT = ".") %>%
  mutate(FEATURE = c("Total", "Deaths", "Severe Injuries", "Light Injuries",
                     rep(c("Total", "Deaths", "Injuries"),3))) %>%
  mutate(CATEGORY = c(replicate(4,"Total"),
                      replicate(3, "Pedestrians"),
                      replicate(3, "Motorcyclists"),
                      replicate(3, "Cyclists"))) %>%
  filter( CATEGORY == 'Total' & FEATURE != 'Total') %>%
  ggplot(aes(x=factor(FEATURE, levels = c("Light Injuries", "Severe Injuries",
                                          "Deaths")),y=COUNT,fill=FEATURE))+
  geom_bar(stat='identity') +
  geom_text(aes(label=comma(COUNT)), 
            position=position_dodge(width=.2),vjust=-2) +
  scale_y_continuous(expand = expansion(add=c(0,6000))) +
  labs(title="Breakdown by Injury Type", x='', y='') +
  theme(legend.position = 'none', panel.background = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(face='bold', size=10)) 
  

```



**The third** group breaks down the type of vehicles involved in the accident.
As above, we have:  

* NB_VEH_IMPLIQUES_ACCDN: Total number of vehicles involved.
  + nb_automobile_camion_leger: Number of automobiles and lightweight trucks.
  + nb_camionLourd_tractRoutier: Number of heavy trucks and tractors.
  + nb_outil_equipement: Number of Tool vehicles and equipment.
  + nb_tous_autobus_minibus: Number of buses, school buses and minibuses
  + nb_bicyclette: Number of bicycles. 
  + nb_cyclomoteur: Number of scooters.
  + nb_motocyclette: Number of motorcycles.
  + nb_taxi: Number of taxis.
  + nb_urgence: Number of emergency vehicle.
  + nb_motoneige: Number of snowmobiles.
  + nb_VHR: Number of dirt bikes.
  + nb_autres_types: Number of other types of vehicles. 
  + nb_veh_non_precise: Number of vehicles with unknown type.
  
```{r fig.width=12, fig.height=8, warning=FALSE}

  accidents %>%
  select( nb_automobile_camion_leger, nb_camionLourd_tractRoutier,
         nb_outil_equipement, nb_tous_autobus_minibus, nb_bicyclette,
         nb_cyclomoteur, nb_motocyclette, nb_taxi, nb_urgence,
         nb_motoneige, nb_VHR, nb_autres_types, nb_veh_non_precise) %>%
  filter(complete.cases(.)) %>%
  colSums() %>%
  as.data.frame %>%
  rename(COUNT = ".") %>%
  mutate(PERC = COUNT/
           sum(accidents$NB_VEH_IMPLIQUES_ACCDN, na.rm=TRUE)*100) %>%
  mutate(PERC = paste0('(', round(PERC, 3),"%)")) %>%
  mutate(TYPE = c( "Automobiles & Light Trucks", "Heavy Trucks & Tractors",
                  "Equipements", "Buses", "Bicycles", "Scooters", "Motorcycles",
                  "Taxis", "Emergencies", "Snowmobiles", "Dirt Bikes", "Others",
                  "Not Specified")) %>%
  ggplot(aes(x=reorder(TYPE, - COUNT), y=COUNT, fill = TYPE)) +
  geom_bar(stat='identity', alpha=.4,,colour='#00abff') +
  coord_flip() +
  geom_text(aes(label = paste0(comma(COUNT),'  ',PERC)),position=position_dodge(width=.2),
            hjust = -.2) +
  theme(legend.position = 'none', 
        axis.text.y = element_text(face='bold', size=10)) +
  scale_y_continuous(expand = expansion(add=c(0,60000))) +
  labs(title = "Breakdown of the total number of vehicles involved by the type",
       x='', y='') +
  theme(panel.background = element_blank(), axis.ticks.y = element_blank())

```
  
There are 3 rows with null values for all features above.   
    
----   

## 3. Accident nature  

This group concerns the type of accidents and contains two variables:

* CD_GENRE_ACCDN: Type of accident. It has the following values
  + 31: Collision with a vehicle. 
  + 32: Collision with a pedestrian.
  + 33: Collision with a cyclist.
  + 34: Collision with a train.
  + 35: Collision with a deer.
  + 36: Collision with a moose, bear, or a caribou.
  + 37: Collision with another animal.
  + 38: Collision with a temporary obstacle.
  + 39: Collision with a detached object (either from a vehicle or a construction).
  + 40: Collision with a street lamp.
  + 41: Collision with a traffic light.
  + 42: Collision with a public pole.
  + 43: Collision with a tree.
  + 44: Collision with a guardrail.
  + ....
  + ....
  + ....
  +   : Not specified (4.9% of the data).
* CD_SIT_PRTCE_ACCDN: Describes a particular situation for when an accident happens.
Mostly null values (97.7%).
  + 1: Spill.
  + 2: Load loss.
  + 3: Snow removal car is involved.
  + 9: Other particular situation causing the accident.
  +  : Not specified. (97.7%)
  


----   
    
## 4. Location variables  

**Finally**, we look at the location and street conditions. 
There are 25 features, however, we have chosen a relevant subset.  

* VITESSE_AUTOR: Authorized speed in KM/H. (39.4% not specified).

```{r fig.width=12, fig.height=8, warning=FALSE}
accidents  %>%
  select(VITESSE_AUTOR, AN) %>%
  group_by(AN) %>%
  table %>%
  as.data.frame %>%
  ggplot(aes(x=factor(VITESSE_AUTOR), y=Freq, fill=VITESSE_AUTOR)) +
  geom_bar(stat='identity') +
  facet_wrap(. ~ factor(AN), nrow = 3) +
  labs(title = "Authorised speed per year for when the accidents happenned",
       x = 'Authorised Speed', y ='Frequency of Accidents')+
  theme(legend.position = 'none', panel.background = element_blank())
#------------------------------------------------------------------------ 
```

* CD_ETAT_SURF: Surface condition:
  + 11: Dry (No spill)
  + 12: Wet
  + 13: Water accumulation
  + 14: Sand or gravel
  + 15: Slush
  + 16: Snow
  + 17: Hard snow
  + 18: Ice
  + 19: Muddy
  + 20: Oily
  + 99: Other (0.3%)
  +   : Not specified (6.0%)

  

```{r fig.width=12, fig.height=8, warning=FALSE}
accidents %>%
  select(CD_ETAT_SURFC) %>%
  replace(is.na(.), 0) %>%
  table(useNA = 'ifany') %>%
  as.data.frame %>%
  mutate(CD_ETAT_SURFC = factor(CD_ETAT_SURFC,
                                levels=c(11,12,13,14,15,16,17,
                                         18,19,20,99,0),
                                labels = c('Dry', 'Wet', 'Water accumulation',
                                           'Sand or gravel','Slush','Snow',
                                           'Hard snow','Icy','Muddy','Oily',
                                           'Other','Not specified'))) %>%
  mutate(PERC = Freq / nrow(accidents) * 100) %>%
  mutate(PERC = paste0('(', round(PERC, 2),"%)")) %>%
  ggplot(aes(x=reorder(CD_ETAT_SURFC,-Freq), y= Freq, fill=CD_ETAT_SURFC)) +
  geom_bar(stat='identity', alpha=.4,colour='#00abff') +
  coord_flip() +
  geom_text(aes(label = paste0(comma(Freq),'  ',PERC)),position=position_dodge(width=.2),
            hjust = -.2) +
  theme(legend.position = 'none', 
        axis.text.y = element_text(face='bold', size=10)) +
  scale_y_continuous(expand = expansion(add=c(0,60000))) +
  labs(title = "Frequency of accidents per surface type",
       x='', y='') +
  theme(panel.background = element_blank(), axis.ticks.y = element_blank())
#-----------------------------------------------------------------------------
```

* CD_ECLRM: Lightning at the moment of the accident:
  + 1: Day and clear
  + 2: Day and semi-dark
  + 3: Night and lighted
  + 4: Night and dark
  +  : Not specified (6.1%)

     
```{r fig.width=12, fig.height=8, warning=FALSE}
accidents %>%
  select(CD_ECLRM) %>%
  replace(is.na(.), 0) %>%
  table(useNA = 'ifany') %>%
  as.data.frame %>%
  mutate(CD_ECLRM = factor(CD_ECLRM, levels = c(1,2,3,4,0),
                           labels =  c('Day and clear', 'Day and semi-dark',
                                       'Night and lighted','Night and dark',
                                       'Not specified' ))) %>%
  mutate(PERC = Freq / nrow(accidents) * 100) %>%
  mutate(PERC = paste0('(', round(PERC, 2),"%)")) %>%
  ggplot(aes(x=reorder(CD_ECLRM,-Freq), y= Freq, fill=CD_ECLRM)) +
  geom_bar(stat='identity', alpha=.4,colour='#00abff') +
  coord_flip() +
  geom_text(aes(label = paste0(comma(Freq),'  ',PERC)),position=position_dodge(width=.2),
            hjust = -.2) +
  theme(legend.position = 'none', 
        axis.text.y = element_text(face='bold', size=10)) +
  scale_y_continuous(expand = expansion(add=c(0,60000))) +
  labs(title = "Lightning at the time of the accident",
       y='Number of Accidents', x='') +
  theme(panel.background = element_blank(), axis.ticks.y = element_blank())
#-------------------------------------------------------------------------
```


* CD_ENVRN_ACCDN: Environment (The dominant activity of the sector where the accident happenned):
  + 1: School area
  + 2: Residential
  + 3: Commercial
  + 4: Industrial
  + 5: Rural
  + 6: Forest
  + 7: Park
  + 9: Other (example: lakes)
  +  : Not specified (3.3%)
     
      
```{r fig.width=12, fig.height=8, warning=FALSE}

accidents %>%
  select(CD_ENVRN_ACCDN) %>%
  replace(is.na(.), 0) %>%
  table(useNA = 'ifany') %>%
  as.data.frame %>%
  mutate(CD_ENVRN_ACCDN = factor(CD_ENVRN_ACCDN, levels = c(1,2,3,4,5,6,7,9,0),
                           labels =  c('School area','Residential','Commercial',
                                       'Industrial','Rural','Forest','Park',
                                       'Other','Not specified' ))) %>%
  mutate(PERC = Freq / nrow(accidents) * 100) %>%
  mutate(PERC = paste0('(', round(PERC, 2),"%)")) %>%
  ggplot(aes(x=reorder(CD_ENVRN_ACCDN,-Freq), y= Freq, fill=CD_ENVRN_ACCDN)) +
  geom_bar(stat='identity', alpha=.4,colour='#00abff') +
  coord_flip() +
  geom_text(aes(label = paste0(comma(Freq),'  ',PERC)),position=position_dodge(width=.2),
            hjust = -.2) +
  theme(legend.position = 'none', 
        axis.text.y = element_text(face='bold', size=10)) +
  scale_y_continuous(expand = expansion(add=c(0,20000))) +
  labs(title = "The domminant activity in the area where accidents happenned",
       y='Number of Accidents', x='') +
  theme(panel.background = element_blank(), axis.ticks.y = element_blank())
#-------------------------------------------------------------------------

```

* CD_ETAT_CHASS: The quality of the road:
  + 1: In a good condition
  + 2: Under construction / repair
  + 3: Ruts (Deformation of the asphalt)
  + 4: Significant cracks
  + 5: Potholes
  + 6: Slopes and different leveling. 
  + 9: Other (0.1%)
  + 0: Not specified. (76.0%)
   
    
```{r fig.width=12, fig.height=8, warning=FALSE}
accidents %>%
  select(CD_ETAT_CHASS) %>%
  replace(is.na(.), 0) %>%
  table(useNA = 'ifany') %>%
  as.data.frame %>%
  mutate(CD_ETAT_CHASS = factor(CD_ETAT_CHASS, levels = c(1,2,3,4,5,6,9,0),
                                 labels =  c('In a good condition',
                                             'Under construction', 'Ruts',
                                             'Significant cracks', 'Potholes',
                                             'Slopes', 'Other',
                                             'Not specified' ))) %>%
  mutate(PERC = Freq / nrow(accidents) * 100) %>%
  mutate(PERC = paste0('(', round(PERC, 2),"%)")) %>%
  ggplot(aes(x=reorder(CD_ETAT_CHASS,-Freq), y= Freq, fill=CD_ETAT_CHASS)) +
  geom_bar(stat='identity', alpha=.4,colour='#00abff') +
  coord_flip() +
  geom_text(aes(label = paste0(comma(Freq),'  ',PERC)),position=position_dodge(width=.2),
            hjust = -.2) +
  theme(legend.position = 'none', 
        axis.text.y = element_text(face='bold', size=10)) +
  scale_y_continuous(expand = expansion(add=c(0,30000))) +
  labs(title = "The Quality of the Road",
       y='Number of Accidents', x='') +
  theme(panel.background = element_blank(), axis.ticks.y = element_blank())
#-----------------------------------------------------------------------------
```

* CD_COND_METEO: Weather conditions:
  + 11: Clear sky
  + 12: Cloudy
  + 13: Foggy
  + 14: Drizzle
  + 15: Heavy rain
  + 16: Strong wind (no blowing snow or rain)
  + 17: Snow / hail
  + 18: Snowstorms and blowing snow
  + 19: Black ice
  + 99: Other (0.6%)
  +   : Not specified (6.4%)
  
     
```{r fig.width=12, fig.height=8, warning=FALSE}
accidents %>%
  select(CD_COND_METEO) %>%
  replace(is.na(.), 0) %>%
  table(useNA = 'ifany') %>%
  as.data.frame %>%
  mutate(CD_COND_METEO = factor(CD_COND_METEO, levels = c(11,12,13,14,15,16,
                                                          17,18,19,99,0),
                                labels =  c('Clear Sky','Cloudy','Foggy','Drizzle',
                                            'Heavy Rain','Strong Wind', 'Snow',
                                            'Snowstorms', 'Black Ice', 'Other',
                                            'Not specified' ))) %>%
  mutate(PERC = Freq / nrow(accidents) * 100) %>%
  mutate(PERC = paste0('(', round(PERC, 2),"%)")) %>%
  ggplot(aes(x=reorder(CD_COND_METEO,-Freq), y= Freq, fill=CD_COND_METEO)) +
  geom_bar(stat='identity', alpha=.4,colour='#00abff') +
  coord_flip() +
  geom_text(aes(label = paste0(comma(Freq),'  ',PERC)),position=position_dodge(width=.2),
            hjust = -.2) +
  theme(legend.position = 'none', 
        axis.text.y = element_text(face='bold', size=10)) +
  scale_y_continuous(expand = expansion(add=c(0,30000))) +
  labs(title = "Weather Conditions",
       y='Number of Accidents', x='') +
  theme(panel.background = element_blank(), axis.ticks.y = element_blank())



```


--------------------------------------


